The Impact of Accretion Disk Winds on the Optical Spectra of Cataclysmic Variables

### 2015-01-15 draft.tex
tweaked figures and text

\begin{figure}	%fullpage
\centering
\includegraphics[width=0.8\textwidth]{figures/fig1.eps}
\caption{Optical spectra of three nova-like variables: 
RW Sex (top; Beuermann et al. 1992),
IX Vel (top middle; Beuermann \& Thomas 1990) 
and RW Tri in and out of eclipse (bottom two panels; Groot et al. 2004).
The data have been digitized from the respective publications and a few of the more
prominent emission and absorption lines are marked. 
These systems have inclinations of $30^\circ$, $60^\circ$ and $70^\circ$ 
respectively. 
The trend of increasing Balmer line emission with inclination can be seen.
In RW Tri strong single-peaked emission in the Balmer lines is seen even
in eclipse, indicating that the lines may be formed in a spatially
extensive disk wind, and there is even a suggestion 
of a (potentially wind-formed) recombination continuum in the eclipsed
spectrum. We have attempted to show each spectrum over a similar dynamic range.
}
\label{novalikes}
\end{figure}

@@

\begin{table}
\centering
\begin{tabular}{p{2cm}p{2cm}p{2cm}}
Model Parameters \\
\hline Parameter 	&	 Model A  & Model B \\ 
\hline \hline 
$M_{WD}$ 	 &	 $0.8 M_{\odot}$  &     \\ 
$R_{WD}$ 	 &	 $7\times10^{8}$cm  & \\ 
$T_{WD}$ 	 &	 $40,000$K        &  \\
$M_{2}$ 	 &	 $0.6 M_{\odot}$  &   \\ 
$\dot{M}_{acc}$ 	 &	 $10^{-8}~M_{\odot}yr^{-1}$  &\\ 
$\dot{M}_{wind}$  &	$10^{-9}~M_{\odot}yr^{-1}$å  & \\ 
$r_{min}$ 	&	 $4 R_{WD}$ &  \\ 
$r_{max}$ 	&	 $12 R_{WD}$  &  \\ 
$\theta_{min}$ 	&	 $20.0^{\circ}$  &  \\ 
$\theta_{max}$ 	&	 $65.0^{\circ}$  &  \\ 
$\gamma$ 	&	 $1$  &  \\ 
$v_{\infty}$ 	&	 $3v_{esc}$  &  \\ 
%%$R_v$ 	        &	 $7\times10^{10}$cm \\ 
$R_v$ 	        &	 $100 R_{WD}$  &  $142.9 R_{WD}$  \\ 
$\alpha$ 	&	 $1.5$   &   $4$\\
\end{tabular}
\centering
\caption{
Parameters used for the geometry and kinematics of the benchmark 
CV model (Model A), which is optimized for the UV band, and a model
which is optimized for the optical band and described in section 6 (Model B).
For Model B, only parameters which are altered are given - otherwise the
Model A parameter is used.
{\bf JM: this needs some stuff adding, as is mentioned in certain places}}
\label{wind_param}
\label{modelb_table}
\end{table}

@@

Our main goal is to test whether the type of disk wind model that has
been successful in explaining the UV spectra of CVs could also have 
significant impact on the optical continuum and emission line spectra
of these systems. In order to set a benchmark, we therefore begin by
investigating one of the fiducial CV wind models that was used by SV93
and LK02 to simulate the UV spectrum of a typical high-state
system. The specific parameters for this model (Model A) are listed in
Table~1. A key point is that the wind mass-loss rate in this model is
set to 10$\%$ of the accretion rate through the disk.

@@

\begin{figure} 
\includegraphics[width=0.45\textwidth]{figures/modela_opt_cont.eps}
\caption{Synthetic optical spectra computed for 
sightlines of 10, 27.5, 45, 62.5 and 80 degrees. In these plots
the flux is divided by a polynomial fit to the 
underlying continuum redward of the Balmer edge, so that 
line-to-continuum ratios and the true depth of the
Balmer jump can be shown.}
\label{spec_continuum}
\end{figure}

\begin{figure} 
\includegraphics[width=0.5\textwidth]{figures/modela_escaping.eps}
\caption{Total packet-binned spectra across all viewing angles. 
The thick black line shows the total 
integrated escaping spectrum, while green and red lines show the contributions from photons that originate in the disk and wind respectively. 
Recombination continuum emission blueward of the Balmer 
edge is already prominent relative to other wind continuum processes, but is not sufficient
to fill in the Balmer jump in this specific model}
\label{cont}
\end{figure}

@@

The benchmark model discussed in section~\ref{modela} was originally
designed to reproduce the wind-formed lines seen in the UV spectra of
high-state CVs. As we have seen, this model does produce significant
amounts of optical emission. However, with only relatively minor
changes to the kinematics of the benchmark model, it is possible to
construct a model that more closely matches the observed optical
spectra of CVs (imaginatively dubbed `Model B').

@@

\begin{figure} 
\includegraphics[width=0.45\textwidth]{figures/modelb_opt_cont.eps}
\caption{
Synthetic optical spectra computed for 
sightlines of 10, 27.5, 45, 62.5 and 80 degrees. In these plots
the flux is divided by a polynomial fit to the 
underlying continuum redward of the Balmer edge, so that 
line-to-continuum ratios and the true depth of the
Balmer jump can be shown.
}
\label{continuumb}
\end{figure}

\begin{figure} 
\includegraphics[width=0.5\textwidth]{figures/modelb_escaping.eps}
\caption{Total packet-binned spectra across all viewing angles. 
The thick black line shows the total 
integrated escaping spectrum, while green and red lines show the contributions from photons that originate in the disk and wind respectively. 
Recombination continuum emission blueward of the Balmer 
edge is already prominent relative to other wind continuum processes, but is not sufficient
to fill in the Balmer jump in this specific model}
\label{modelb_escape}
\end{figure}

@@

There are two possible ways in which the model
could be altered to better reproduce single-peaked
lines as a result of radial velocity shear. First,
one could adopt a more equatorial wind geometry. 
This would increase the velocity gradient in the radial 
direction. However, CV winds are known to be moderately collimated.
This is demonstrated by the fact that blue-shifted resonance line 
absorption is seen in low inclination systems such as RW Sex \citep{greensteinoke1982},
but not in eclipsing systems \citep{cordovamason1985}.
Furthermore, the main reason for adopting this type of 
wind geometry is that it already successfully models the 
P-Cygni resonance lines seen in the UV (LK02, Noebauer et al. 2010),
Equatorial winds are expected in AGN and X-ray binaries 
\citep{weymann1991, ponti2012}, so these winds may be more likely
to possess the required kinematics.
Second, one could adopt velocity laws with higher 
velocity gradients in the line emission region. 
However, this poses problems for a smooth biconical wind. 
We have already demonstrated that high densities are required 
to produce significant emission. In a smooth model, which
satisfies the mass continuity condition (see section 3.1), 
it is therefore difficult to reconcile 
these high densities with the high velocity gradients required.

@@

\begin{figure} %fullpage
\includegraphics[width=0.9\textwidth]{figures/fig13_eclipse.eps}
\caption{{\sl Top Panel:} In and out of eclipse spectra of the high
inclination NL RW Tri. {\sl Bottom Panel:} In and out of eclipse synthetic
spectra from Model B.
The artificial `absorption' feature just redward of the Balmer jump
is caused for the reasons described in section 5.2.}
\label{rwtricomp}
\end{figure} %fullpage

>>>>

\begin{figure}	%fullpage
\centering
\includegraphics[width=0.8\textwidth]{figures/fig1.eps}
% \caption{Optical spectra of three nova-like variables: 
% RW Sex (top; Beuermann et al. 1992),
% IX Vel (top middle; Beuermann \& Thomas 1990) 
% and RW Tri in and out of eclipse (bottom two panels; Groot et al. 2004).
% The data have been digitized from the respective publications and a few of the more
% prominent emission and absorption lines are marked. 
\caption{
Optical spectra of three nova-like variables: 
RW Sex (top; Beuermann et al. 1992),
IX Vel (top middle; A. Pala \& B. Gaensicke, private communication) 
and RW Tri in and out of eclipse (bottom two panels; Groot et al. 2004).
The data for RW Sex and RW Tri were digitized from the respective publications,
and the IX Vel spectrum was obtained using XShooter on the VLT on the night of ??.
These systems have approximate inclinations of $30^\circ$, $60^\circ$ and $70^\circ$ 
(see section 5.4) respectively. 
The trend of increasing Balmer line emission with inclination can be seen.
In RW Tri strong single-peaked emission in the Balmer lines is seen even
in eclipse, indicating that the lines may be formed in a spatially
extensive disk wind, and there is even a suggestion 
of a (potentially wind-formed) recombination continuum in the eclipsed
spectrum. We have attempted to show each spectrum over a similar dynamic range.
}
\label{novalikes}
\end{figure}

@@

\begin{table}
\centering
\begin{tabular}{p{2cm}p{2cm}p{2cm}}
Model Parameters \\
\hline Parameter 	&	 Model A  & Model B \\ 
\hline \hline 
$M_{WD}$ 	 &	 $0.8 M_{\odot}$  &     \\ 
$R_{WD}$ 	 &	 $7\times10^{8}$cm  & \\ 
$T_{WD}$ 	 &	 $40,000$K        &  \\
$M_{2}$ 	 &	 $0.6 M_{\odot}$  &   \\ 
$q$ 	 &	 $0.75 M_{\odot}$  &   \\ 
$P_{orb}$ 	 &	 $5.57$~hr  &   \\ 
$\dot{M}_{acc}$ 	 &	 $10^{-8}~M_{\odot}yr^{-1}$  &\\ 
$\dot{M}_{wind}$  &	$10^{-9}~M_{\odot}yr^{-1}$å  & \\ 
$r_{min}$ 	&	 $4 R_{WD}$ &  \\ 
$r_{max}$ 	&	 $12 R_{WD}$  &  \\ 
$\theta_{min}$ 	&	 $20.0^{\circ}$  &  \\ 
$\theta_{max}$ 	&	 $65.0^{\circ}$  &  \\ 
$\gamma$ 	&	 $1$  &  \\ 
$v_{\infty}$ 	&	 $3v_{esc}$  &  \\ 
%%$R_v$ 	        &	 $7\times10^{10}$cm \\ 
$R_v$ 	        &	 $100 R_{WD}$  &  $142.9 R_{WD}$  \\ 
$\alpha$ 	&	 $1.5$   &   $4$\\
\end{tabular}
\centering
\caption{
Parameters used for the geometry and kinematics of the benchmark 
CV model (model A), which is optimized for the UV band, and a model
which is optimized for the optical band and described in section 5 (model B).
For model B, only parameters which are altered are given - otherwise the
model A parameter is used.
{\bf JM: this needs some stuff adding, as is mentioned in certain places}}
\label{wind_param}
\label{modelb_table}
\end{table}

@@

Our main goal is to test whether the type of disk wind model that has
been successful in explaining the UV spectra of CVs could also have 
significant impact on the optical continuum and emission line spectra
of these systems. In order to set a benchmark, we therefore begin by
investigating one of the fiducial CV wind models that was used by SV93
and LK02 to simulate the UV spectrum of a typical high-state
system. The specific parameters for this model (model A) are listed in
Table~1. A key point is that the wind mass-loss rate in this model is
set to 10$\%$ of the accretion rate through the disk.

@@

\begin{figure} 
\includegraphics[width=0.45\textwidth]{figures/modela_opt_cont.eps}
\caption{Synthetic optical spectra from model A computed for 
sightlines of 10, 27.5, 45, 62.5 and 80 degrees. In these plots
the flux is divided by a polynomial fit to the 
underlying continuum redward of the Balmer edge, so that 
line-to-continuum ratios and the true depth of the
Balmer jump can be shown.}
\label{spec_continuum}
\end{figure}

\begin{figure} 
\includegraphics[width=0.5\textwidth]{figures/modela_escaping.eps}
\caption{Total packet-binned spectra across all viewing angles. 
The thick black line shows the total 
integrated escaping spectrum, 
while the green line shows disk photons which escape without being reprocessed by
the wind. The red line show the contributions from reprocessed photons. 
Recombination continuum emission blueward of the Balmer 
edge is already prominent relative to other wind continuum processes, but is not sufficient
to fill in the Balmer jump in this specific model}
\label{cont}
\end{figure}

@@

The benchmark model discussed in section~\ref{modela} was originally
designed to reproduce the wind-formed lines seen in the UV spectra of
high-state CVs. As we have seen, this model does produce significant
amounts of optical emission. However, with only relatively minor
changes to the kinematics of the benchmark model, it is possible to
construct a model that more closely matches the observed optical
spectra of CVs.

@@

\begin{figure} 
\includegraphics[width=0.45\textwidth]{figures/modelb_opt_cont.eps}
\caption{
Synthetic optical spectra from model B computed for 
sightlines of 10, 27.5, 45, 62.5 and 80 degrees. In these plots
the flux is divided by a polynomial fit to the 
underlying continuum redward of the Balmer edge, so that 
line-to-continuum ratios and the true depth of the
Balmer jump can be shown.
}
\label{continuumb}
\end{figure}

\begin{figure} 
\includegraphics[width=0.5\textwidth]{figures/modelb_escaping.eps}
\caption{Total packet-binned spectra across all viewing angles. 
The thick black line shows the total 
integrated escaping spectrum, 
while the green line shows disk photons which escape without being reprocessed by
the wind. The red line show the contributions from reprocessed 
photons. 
In this denser model the reprocessed contribution is significant compared
to the escaping disk spectrum. The Balmer continuum emission is prominent, and
we can see a clear effect on the overall spectral shape from the wind.}
\label{modelb_escape}
\end{figure}

@@

There are two possible ways in which the model
could be altered to better reproduce single-peaked
lines as a result of radial velocity shear. First,
one could adopt a more equatorial wind geometry. 
This would increase the velocity gradient in the radial 
direction. However, CV winds are known to be moderately collimated.
This is demonstrated by the fact that blue-shifted resonance line 
absorption is seen in low inclination systems such as RW Sex \citep{greensteinoke1982},
but not in eclipsing systems \citep{cordovamason1985}.
Furthermore, the main reason for adopting this type of 
wind geometry is that it already successfully models the 
P-Cygni resonance lines seen in the UV (LK02; Noebauer et al. 2010),
Equatorial winds are expected in AGN and X-ray binaries 
\citep{weymann1991, ponti2012}, so these winds may be more likely
to possess the required kinematics.
Second, one could adopt velocity laws with higher 
velocity gradients in the line emission region. 
However, this poses problems for a smooth biconical wind. 
We have already demonstrated that high densities are required 
to produce significant emission. In a smooth model, which
satisfies the mass continuity condition (see section 3.1), 
it is therefore difficult to reconcile 
these high densities with the high velocity gradients required.

@@

\begin{figure} %fullpage
\includegraphics[width=0.9\textwidth]{figures/fig13_eclipse.eps}
\caption{{\sl Top Panel:} In and out of eclipse spectra of the high
inclination NL RW Tri. {\sl Bottom Panel:} In and out of eclipse synthetic
spectra from model B.
The artificial `absorption' feature just redward of the Balmer jump
is caused for the reasons described in section 5.2.}
\label{rwtricomp}
\end{figure} %fullpage
